---
title: "Introduction to Cardinal"
author: "Kylie Ariel Bemis"
date: "Revised: October 27, 2019"
output:
  BiocStyle::html_document:
  toc: true
vignette: >
  %\VignetteIndexEntry{Basics: Introduction to Cardinal}
  %\VignetteKeyword{ExperimentData, MassSpectrometryData, ImagingMassSpectrometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo=FALSE, results='asis'}
BiocStyle::markdown()
```

```{r setup, echo=FALSE, message=FALSE}
library(Cardinal)
register(SerialParam())
options(Cardinal.verbose=FALSE)
options(Cardinal.progress=FALSE)
RNGkind("L'Ecuyer-CMRG")
```

# Introduction 

# Importing MS imaging data from imzML

## "Continuous" versus "Processed" imzML

## Tips and tricks

# Components of an `MSImagingExperiment`

## Pixel metadata

## Feature metadata

## Accessing spectra data

## Subsetting and combining MSI datasets

## Tips and tricks


# Visualization

## Plotting mass spectra

## Plotting ion images

## Region-of-interest selection

## Saving plots and images

## Plotting options


# Manipulation and summarization

## Transforming and filtering data

## Summarizing an MSI dataset


# Useful R functions to know


# Constructing an `MSImagingExperiment` from scrach


# Session information

```{r session-info}
sessionInfo()
```



