---
title: "Segmentation of MSI experiments with Cardinal"
author: "Kylie Ariel Bemis"
date: "Revised: October 27, 2019"
output:
  BiocStyle::html_document:
  toc: true
vignette: >
  %\VignetteIndexEntry{Segmentation: Unsupervised analysis workflow}
  %\VignetteKeyword{ExperimentData, MassSpectrometryData, ImagingMassSpectrometry, Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r style, echo=FALSE, results='asis'}
BiocStyle::markdown()
```

```{r setup, echo=FALSE, message=FALSE}
library(CardinalWorkflows)
register(SerialParam())
options(Cardinal.verbose=FALSE)
options(Cardinal.progress=FALSE)
RNGkind("L'Ecuyer-CMRG")
```

# Introduction 

# Segmentation of a pig fetus wholy body cross section

## Pre-processing

## Visualization

### Ion images

### Prinipal components analysis (PCA)

## Segmentation with spatial shrunken centroids (SSC)

### Plotting the segmentation

### Plotting the (shrunken) mean spectra

### Plotting and interpretting t-statistics of the *m/z* values

### Selecting a segmentation

### Interpretting the segmentation

# Segmentation of a cardinal painting

## Pre-processing

## Visualization

## Segmetation with SSC


# Session information

```{r session-info}
sessionInfo()
```



