---
title: "Classification of MSI experiments with Cardinal"
author: "Kylie Ariel Bemis"
date: "Revised: October 27, 2019"
output:
  BiocStyle::html_document:
  toc: true
vignette: >
  %\VignetteIndexEntry{Classification: Supervised analysis workflow}
  %\VignetteKeyword{ExperimentData, MassSpectrometryData, ImagingMassSpectrometry, Classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
  ```{r style, echo=FALSE, results='asis'}
BiocStyle::markdown()
```

```{r setup, echo=FALSE, message=FALSE}
library(Cardinal)
register(SerialParam())
options(Cardinal.verbose=FALSE)
options(Cardinal.progress=FALSE)
RNGkind("L'Ecuyer-CMRG")
```

# Introduction 

# Analysis of a renal cell carcinoma (RCC) cancer dataset

## Pre-processing

## Visualization

### Ion images

### Prinipal components analysis (PCA)

## Classification with spatial shrunken centroids (SSC)

### Cross-validation with SSC

### Plotting the classified images

### Plotting the (shrunken) mean spectra

### Plotting and interpretting t-statistics of the *m/z* values

# Additional notes on cross-validation


# Session information

```{r session-info}
sessionInfo()
```



