---
title: "Class comparison for MSI experiments with Cardinal"
author: "Kylie Ariel Bemis"
date: "Revised: October 27, 2019"
output:
  BiocStyle::html_document:
  toc: true
vignette: >
  %\VignetteIndexEntry{Class comparision: Statistical testing workflow}
  %\VignetteKeyword{ExperimentData, MassSpectrometryData, ImagingMassSpectrometry, Regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
  ```{r style, echo=FALSE, results='asis'}
BiocStyle::markdown()
```

```{r setup, echo=FALSE, message=FALSE}
library(CardinalWorkflows)
register(SerialParam())
options(Cardinal.verbose=FALSE)
options(Cardinal.progress=FALSE)
RNGkind("L'Ecuyer-CMRG")
```

# Introduction 

# Statistical testing for a renal cell carcinoma (RCC) cancer dataset

## Pre-processing

### Non-specific filtering to reduce data size

### Segmentation with spatial Dirichlet Gaussian mixture model (DGMM)

## Visualization

## Class comparison with means-based testing

### Fitting models with means-summarized groups

### Interpreting the results

## Class comparison with segmentation-based testing

### Fitting models with spatial-DGMM-summarized groups

### Interpreting the results

# Additional uses for spatial DGMM segmentations

# Session information

```{r session-info}
sessionInfo()
```



